---
title: "Lab_4_markdown"
output: html_document
date: "2024-09-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
####----------------------------------------------------------#
#### Preamble
####----------------------------------------------------------#

# Load libraries.
## install via `install.packages("name")`
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(kableExtra)
library(maps)
library(RColorBrewer)
library(sf)
library(tidyverse)
library(viridis)
```

```{r}
####----------------------------------------------------------#
#### Read, merge, and process data.
####----------------------------------------------------------#

# Read incumbency/vote data.
d_vote <- read_csv("popvote_1948-2020.csv")
d_state_vote <- read_csv("state_popvote_1948_2020.csv")
d_vote$party[d_vote$party == "democrat"] <- "DEM"
d_vote$party[d_vote$party == "republican"] <- "REP"

# Read economic data.
d_econ <- read_csv("fred_econ.csv") |> 
  filter(quarter == 2)

# Read polling and election results data. 
d_pollav_natl <- read_csv("national_polls_1968-2024.csv")
d_pollav_state <- read_csv("state_polls_1968-2024.csv")
```

```{r}
# Shape and merge polling and election data using November polls. 
d_poll_nov <- d_vote |> 
  left_join(d_pollav_natl |> 
              group_by(year, party) |> 
              top_n(1, poll_date) |> 
              select(-candidate), 
            by = c("year", "party")) |> 
  rename(nov_poll = poll_support) |> 
  filter(year <= 2020) |> 
  drop_na()

# Create dataset of polling average by week until the election. 
d_poll_weeks <- d_pollav_natl |> 
  group_by(year, party, weeks_left) |>
  summarize(mean_poll_week = mean(poll_support)) |> 
  filter(weeks_left <= 30) |> 
  pivot_wider(names_from = weeks_left, values_from = mean_poll_week) |> 
  left_join(d_vote, by = c("year", "party"))
```

```{r}
####----------------------------------------------------------#
#### Descriptive statistics on the incumbency advantage. 
####----------------------------------------------------------#

# How many post-war elections (18 and 2024) have there been 
# where an incumbent president won? 
d_vote |> 
  filter(winner) |> 
  select(year, win_party = party, win_cand = candidate) |> 
  mutate(win_party_last = lag(win_party, order_by = year),
         win_cand_last = lag(win_cand, order_by = year)) |> 
  mutate(reelect_president = win_cand_last == win_cand) |> 
  filter(year > 1948 & year < 2024) |> 
  group_by(reelect_president) |> 
  summarize(N = n()) |> 
  mutate(Percent = round(N/sum(N) * 100, 2)) |>
  as.data.frame()
```

```{r}
# A different way of assessing the incumbency advantage 
# (out of 11 elections where there was at least one incumbent running). 
inc_tab <- d_vote |> 
  filter(year > 1948 & year < 2024) |>
  select(year, party, candidate, incumbent, winner) |> 
  pivot_wider(names_from = party, 
              values_from = c(candidate, incumbent, winner)) |> 
  filter(incumbent_DEM == TRUE | incumbent_REP == TRUE)


cat(paste0("Elections with At Least One Incumbent Running: ", nrow(inc_tab), "\n",
   "Incumbent Victories: ", (sum(inc_tab$incumbent_REP & inc_tab$winner_REP) + 
                             sum(inc_tab$incumbent_DEM & inc_tab$winner_DEM)), "\n",
    "Percentage: ", round((sum(inc_tab$incumbent_REP & inc_tab$winner_REP) + 
                           sum(inc_tab$incumbent_DEM & inc_tab$winner_DEM))/
                           nrow(inc_tab)*100, 2)))

```

```{r}
# In the six elections since 2000?
inc_tab |> 
  filter(year >= 2000)
       
# How many post-war elections have there been where the incumbent *party* won? 
d_vote |> 
  filter(winner) |> 
  select(year, win_party = party, win_cand = candidate) |> 
  mutate(win_party_last = lag(win_party, order_by = year),
         win_cand_last = lag(win_cand, order_by = year)) |> 
  mutate(reelect_party = win_party_last == win_party) |> 
  filter(year > 1948 & year < 2024) |> 
  group_by(reelect_party) |> 
  summarize(N = n()) |> 
  mutate(Percent = round(N/sum(N) * 100, 2)) |>
  as.data.frame()

```

```{r}
# How many post-war elections have there been where winner served in 
# previous administration?
100*round(prop.table(table(`prev_admin` = d_vote$prev_admin[d_vote$year > 1948 & 
                                     d_vote$year < 2024 & 
                                     d_vote$winner == TRUE])), 4)
```

```{r}
####----------------------------------------------------------#
#### Pork analysis. 
####----------------------------------------------------------#
 
# Read federal grants dataset from Kriner & Reeves (2008). 
d_pork_state <- read_csv("fedgrants_bystate_1988-2008.csv")

# What strategy do presidents pursue? 
d_pork_state |> 
  filter(!is.na(state_year_type)) |> 
  group_by(state_year_type) |>
  summarize(mean_grant = mean(grant_mil, na.rm = T), se_grant = sd(grant_mil, na.rm = T)/sqrt(n())) |> 
  ggplot(aes(x = state_year_type, y = mean_grant, ymin = mean_grant-1.96*se_grant, ymax = mean_grant+1.96*se_grant)) + 
  coord_flip() + 
  geom_bar(stat = "identity", fill = "chartreuse4") + 
  geom_errorbar(width = 0.2) + 
  labs(x = "Type of State & Year", 
       y = "Federal Grant Spending (Millions of $)", 
       title = "Federal Grant Spending (Millions $) by State Election Type") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
```
core state is with a larger number of co-partisans 
not significant within core or swing, between the two, swing much more

```{r}
# Do presidents strategize for their successor as well? 
d_pork_state |> 
  filter(!is.na(state_year_type2)) |> 
  group_by(state_year_type2) |>
  summarize(mean_grant = mean(grant_mil, na.rm = T), se_grant = sd(grant_mil, na.rm = T)/sqrt(n())) |> 
  ggplot(aes(x = state_year_type2, y = mean_grant, ymin = mean_grant-1.96*se_grant, ymax = mean_grant+1.96*se_grant)) + 
  coord_flip() + 
  geom_bar(stat = "identity", fill = "chartreuse4") + 
  geom_errorbar(width = 0.2) + 
  labs(x = "Type of State & Year", 
       y = "Federal Grant Spending (Millions of $)", 
       title = "Federal Grant Spending (Millions $) by State Election Type") + 
  theme_minimal() +
  theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))

```
what happens when a new party candidate is running, they are not running 
not significantly correct, can't reject the null

```{r}
# Pork county model. 
d_pork_county <- read_csv("fedgrants_bycounty_1988-2008.csv")

pork_mod_county_1 <- lm(dvoteswing_inc  ~ dpct_grants*comp_state + as.factor(year), 
                      d_pork_county)
summary(pork_mod_county_1)

pork_mod_county_2 <- lm(dvoteswing_inc ~ dpct_grants*comp_state + as.factor(year) +
                          dpc_income + inc_ad_diff + inc_campaign_diff + 
                          dhousevote_inc + iraq_cas2004 + iraq_cas2008 + 
                          dpct_popl,
                        data = d_pork_county)
summary(pork_mod_county_2)

```
make incumbent vote swing a dependent variable as a function of the interaction between federal grant spending and state, include uear fixed effects 

```{r}
# Pork state model. 
d_pork_state_model <- d_state_vote |>
  mutate(state_abb = state.abb[match(d_state_vote$state, state.name)]) |>
  inner_join(d_pork_state, by = c("year", "state_abb")) |>
  left_join(d_vote, by = "year") |>
  filter(incumbent_party == TRUE) |>
  mutate(inc_pv2p = ifelse(party == "REP", R_pv2p, D_pv2p)) |>
  mutate(is_comp = case_when(state_year_type == "swing + election year" ~ 1,
                             .default = 0)) |>
  group_by(state) |>
  mutate(change_grant_mil = (1-grant_mil/(lag(grant_mil, n = 1)))*100,
         change_inc_pv2p = (1-inc_pv2p/(lag(inc_pv2p, n = 1)))*100) |>
  ungroup() |>
  select(state, year, is_comp, change_grant_mil, change_inc_pv2p)

pork_state_mod <- lm(change_inc_pv2p ~ is_comp*change_grant_mil + as.factor(year),
                     data = d_pork_state_model)
summary(pork_state_mod)
```
didn't work well at state level
also question on if relevant in 2024 because data ends in 2008 and is better at the county level and we are doing a state level model 
recommend not do for the blogs

```{r}
####----------------------------------------------------------#
#### Time for a change model. 
####----------------------------------------------------------#

# Join data for time for change model.
d_tfc_train <- d_vote |> 
  left_join(d_econ, by = "year") |> 
  filter(incumbent_party) |>
  mutate(incumbent = as.numeric(incumbent))

# Estimate time for change model through 2016.
tfc_mod_2016 <- lm(pv2p ~ GDP_growth_quarterly + incumbent + juneapp, 
                   data = subset(d_tfc_train, year < 2020))
summary(tfc_mod_2016)

# Estimate simplified time for change model for 2020. 
# https://www-cambridge-org.ezp-prod1.hul.harvard.edu/core/services/aop-cambridge-core/content/view/47BBC0D5A2B7913DBB37FDA0542FD7E8/S1049096520001389a.pdf/its-the-pandemic-stupid-a-simplified-model-for-forecasting-the-2020-presidential-election.pdf
tfc_mod_2020 <- lm(pv2p ~ juneapp, 
                   data = subset(d_tfc_train, year < 2024))
summary(tfc_mod_2020)
```

