---
title: 'Week 8: Sudden Shocks and Shoring Up My Model'
author: Jacqui Schlesinger
date: '2024-10-27'
slug: week-8
categories: []
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
set.seed(02138)
```

```{r}
####----------------------------------------------------------#
#### Load Libraries
####----------------------------------------------------------#

library(geofacet)
library(ggpubr)
library(ggthemes)
library(haven)
library(kableExtra)
library(maps)
library(mgcv)
library(mgcViz)
library(RColorBrewer)
library(scales)
library(sf)
library(spData)
library(stargazer)
library(tidygeocoder)
library(tidyverse)
library(tigris)
library(tmap)
library(tmaptools)
library(viridis)
library(ggplot2)
library(plotly)
library(ggrepel)
library(car)
library(purrr)
library(broom)
library(knitr)
library(dplyr)
library(tidyr)
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(tidyverse)
library(knitr)
library(kableExtra)
library(plotly)
library(lubridate)

```

In my last post before my final prediction, I will be exploring the final variables I must evaluate for inclusion in my model-- those related to sudden shocks-- and reviewing professional forecasts and what they have to say about the coming election. Additionally, I will tweak my binomial model specification, and test the in sample fit of my binomial and linear regression models. 

### Review of Professional Forecasts
came out recently
from slides

### Review of Sudden Shocks
data viz
what are they
why I won't use them
lit inclusion

### Tweaking the Binomial Model Specification


### Evaluating the Fit of Binomial and Linear Regression Models



### The Key: Turnout
this new model will be updated in my final prediction, stay tuned

### Data Sources

