---
title: 'Week 1: Introduction and Past National Election Data'
author: Jacqui Schlesinger
date: '2024-09-05'
slug: week-1-introductory-post
categories: []
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, results = 'hide'}
# Load libraries
library(ggplot2)
library(maps)
library(tidyverse)
library(plotly)
library(ggrepel)

```

```{r}
####----------------------------------------------------------#
#### Data read in and cleaning
####----------------------------------------------------------#

### Popular vote data
# Read presidential popular vote
d_popvote <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/popvote_1948-2020.csv") # if you run this code please change to location of data

# Pivot data to wide format with party names as columns and two-party vote share as values.
d_popvote_wide <- d_popvote |>
    select(year, party, pv2p) |>
    pivot_wider(names_from = party, values_from = pv2p)

# Modify winner column to show "D" if Democrats win and "R" if Republicans win
d_popvote_wide <- d_popvote_wide |> 
    mutate(winner = case_when(democrat > republican ~ "D",
                              TRUE ~ "R"))

### Shape file data
# Sequester shapefile of states from `maps` library
states_map <- map_data("state")

### Candidate votes wide data
# Read wide version of dataset that can be used to compare candidate votes with one another
d_pvstate_wide <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/clean_wide_state_2pv_1948_2020.csv") # if you run this code please change to location of data to fit your set up

# Merge d_pvstate_wide with state_map
d_pvstate_wide$region <- tolower(d_pvstate_wide$state)

pv_map <- d_pvstate_wide |>
  left_join(states_map, by = "region")

### Electoral votes data by state and year
ec <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/ec_full.csv") # if you run this code please change to location of data to fit your set up

# Update to handle three years where DC electors is NA (2012, 2016, 2020)
ec <- ec %>%
  mutate(electors = ifelse(state == "District Of Columbia" & is.na(electors), 3, electors))

# note: ec dataset is missing 1960 during which the electoral votes were different than in 1956 or 1964 by state, so I have not manually entered them

### Electoral votes by year and party
# because our data does not include info about third parties, I created this csv so that a line plot of electoral college votes will be appropriate for graphing
ec_yearly <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/electoral_by_year.csv") # if you run this code please change to location of data to fit your set up

```

```{r}
####----------------------------------------------------------#
#### Extension 1 Partial: Create a Custom Theme
####----------------------------------------------------------#

# Create custom theme for maps 
maps_theme <- theme_bw() + 
    theme(panel.border = element_blank(),     
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          plot.title = element_text(size = 12),
          strip.text = element_text(size = 9),
          legend.position = "right",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9))

# Create custom theme for line plots
line_plot_theme <- theme_bw() + 
    theme(panel.border = element_blank(),
          plot.title = element_text(size = 15, hjust = 0.5), 
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(size = 12),
          strip.text = element_text(size = 18),
          axis.line = element_line(colour = "black"),
          legend.position = "bottom",
          legend.text = element_text(size = 12))
```

_In this blog, I will attempt to forecast the outcome of the 2024 US presidential election. As a part of Gov 1347: Election Analytics, I will use data from a variety of sources to develop a compelling model._

This week's focus is on data from national presidential election outcomes, 1948-2020. Before analyzing the factors leading to a candidate's victory, it is important to understand past election trends. The objective of this post is to answer two questions in furtherance of this goal: **how competitive are US presidential elections**, and **which states vote blue/red and how consistently**. To do so, I'll examine Electoral College and two-way popular vote outcomes and implement a simplified Helmut Norpoth model for a baseline prediction solely from prior election results.

### Popular Vote Over Time
Instead of using national popular vote, a more focused analysis looks at the two-way popular vote, considering only votes cast for the two main parties. I first evaluate how the popular vote distribution has changed between the parties.

```{r}
####----------------------------------------------------------#
#### Past Two Way Popular Vote Results
####----------------------------------------------------------#

# Create line plot of R and D two party national popular vote by year
popvote_line <- d_popvote %>%
  ggplot(aes(x = year, y = pv2p, color = party, group=1, text = paste("Year:", year, "<br>2-Way Popular Vote:", round(pv2p, 1), "%"))) + 
  geom_line(stat = "identity") +
  geom_point() +
  scale_color_manual(values = c("dodgerblue4","firebrick1")) +
  xlab("") +
  ylab("Two-Party National Popular Vote (%)") +
  ggtitle("Presidential Vote Share (1948-2020)") + 
  scale_x_continuous(breaks = seq(1948, 2020, 4)) +
  line_plot_theme

ggplotly(popvote_line, tooltip = "text")
```

```{r, results = 'hide'}
####----------------------------------------------------------#
#### Calcs for Paragraph
####----------------------------------------------------------#
diff_calcs <- d_popvote %>%
  group_by(year) %>%
  summarize(
    R_pv2p = pv2p[party == 'republican'],
    D_pv2p = pv2p[party == 'democrat'],
    diff = abs(R_pv2p - D_pv2p)
  ) %>%
  ungroup() %>%
  arrange(desc(diff))

diff_calcs
```

From 1948 to 2020, power shifts between Democrats and Republicans occurred every few election cycles. The margins varied significantly, with some elections resulting in large gaps, like the 23.6 percentage-point difference in 1972 between Nixon and McGovern, and others with extremely tight races, like the 0.17 percentage-point margin in 1960 between Nixon and Kennedy. Over the decades, landslide victories have seemingly become less common, with margins still fluctuating but within consistently tighter races. 

### Electoral College Over Time
The Electoral College also reflects the increased competitiveness seen in popular vote trends. As the true presidential elections determinant, it shows how close they truly were.

```{r}
####----------------------------------------------------------#
#### Past Electoral College Results
####----------------------------------------------------------#

# Create line plot of R and D electoral votes by year, removing other candidates
ecvote_line <- ec_yearly %>%
  filter(party != "Other") %>%
  ggplot(aes(x = year, y = electoral_votes, color = party, group=1, text = paste("Year:", year, "<br>Electors:", electoral_votes))) + 
  geom_line(stat = "identity") +
  geom_point() +
  geom_hline(yintercept = 270, linetype = "dotted", color = "yellow") +
  scale_color_manual(values = c("dodgerblue4","firebrick1")) +
  xlab("") +
  ylab("Electoral College Votes (Number of Electors)") +
  ggtitle("Presidential Electoral College Share (1948-2020)") + 
  scale_x_continuous(breaks = seq(1948, 2020, 4)) +
  line_plot_theme

ggplotly(ecvote_line, tooltip = "text")

```

The dominance of single parties the Electoral College has similarly declined, with fewer recent landslide victories. This points to an _increased competitiveness in presidential elections_.

### Electoral College Versus Two-Way Popular Vote Share
Close elections often hinge on [a few key states](https://books.google.com/books?hl=en&lr=&id=1it4EAAAQBAJ&oi=fnd&pg=PA339&dq=swing+states+became+critical+to+elections&ots=Udo10ht0-f&sig=jVlcrKHiERO7ckvIGSgCJAb4poY#v=onepage&q&f=false) in the Electoral College. Later, I investigate trends within swing states. For now, I examine the national electoral votes and two-way popular votes since 1948, comparing the two margins of victory for elections with the largest and smallest margins:

_1972 (Nixon vs. McGovern)_: Nixon won by margin of 23.6 pp and 503 electoral votes.

_1964 (Goldwater vs. Johnson)_: Johnson won by margin of 22.7 pp and 434 electoral votes.

_1960 (Nixon vs. Kennedy)_: Kennedy won by margin of 0.17 pp and 84 electoral votes.

_2000 (Bush vs. Gore)_: Bush lost the two-way popular vote by 0.5 pp but won electoral college by 5 votes.

As in 2000 (and 2016), fluctuations in both electoral and popular vote results are of similar magnitude, but may not lead to the same outcome. _These cases demonstrate how the national popular vote and Electoral College results don’t always align, underscoring the importance of state-level popular vote outcomes in predicting elections._

### Swing State and Margins Analysis, 1948-2020
To better understand how the popular vote and win margins can effect election outcomes, it’s crucial to analyze changes at the state level. Looking at each state's swing-- the change in the popular vote margin relative to the results in that state from the last election-- provides insight into where elections might be won or lost.

```{r}
####----------------------------------------------------------#
#### Extension 3: Swing State Map Extension
####----------------------------------------------------------#

# Calculate swing map margins for each state and year (D sub y minus D sub y-4) 
pv2p_2024_states <- d_pvstate_wide %>%
  group_by(state) %>%
  mutate(pv2p_margin = D_pv2p - D_pv2p_lag1, region = tolower(state))

# Join data with states shape file and graph swing for years 1980-2020 using facets
pv2p_2024_states %>% 
  filter(year >= 1980) %>%
  left_join(states_map, by = "region") %>%
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pv2p_margin), color = "black") +
  facet_wrap(facets = year ~.) + # 
  ggtitle("Change in Vote Margin From Past Election, 1980-2020") + 
  scale_fill_gradient2(
    high = "dodgerblue4",
    mid = "white",
    low = "firebrick1",
    limits = c(-15, 15),
    name = "Percentage Point Swing (pp)"
  ) +
  maps_theme

```

States shaded darker swung more toward Republicans (red) and Democrats (blue) compared to their margins in the past election. Importantly, dark shading does not mean a state is competitive--it is often the opposite. 

For instance, Utah, consistently Republican, saw a red swing in 2012 when Mitt Romney, [a son of Utah](https://link.springer.com/chapter/10.1057/9781137360823_5), was the candidate. This made 2016 appear blue due to a smaller Republican margin, but Utah hasn't voted Democrat since the landslide in 1964, so it is not a swing state.

Conversely, states with consistently smaller swings, like Pennsylvania and Arizona, are more likely to be true swing states. A swing map visualizes only how states move relatively. For example, in 2008, many states swung more Democratic than the past for Obama, but in 2012, the swing shifted red. However, Obama still won re-election, demonstrating that while swing maps show relative support movement, the popular vote within each state remains crucial.

### The Helmut Norpoth Model
While swings and margins provide valuable insight, they alone cannot predict outcomes. The same goes for the popular vote nationally, which does not always determine electoral votes. But looking at state-level popular vote can provide an early 2024 election forecast based only on past results using a simplified version of the [Helmut Northpoth electoral cycle model](http://primarymodel.com/). This is defined by 

$$ vote_{2024} = 0.75vote_{2020} + 0.25vote_{2016}$$
This is a _predictive model_ of the election, using inputs of past election results to predict future outcomes. The map generated by this model shows predicted margins of victory by state, as well as projected winner and electoral votes.

```{r}
####----------------------------------------------------------#
#### 2024 Margins Forecast
####----------------------------------------------------------#

# Create prediction (pv2p and margin) based on Helmut Norpoth electoral cycle model
# vote_2024 = 3/4*vote_2020 + 1/4*vote_2016 (lag1, lag2, respectively)

# Calculate 2024 pv2p and margin
pv2p_2024_states_margins <- d_pvstate_wide %>%
  filter(year == 2020) %>%
  group_by(state) %>%
  summarize(D_pv2p_2024 = 3/4 * D_pv2p + 1/4 * D_pv2p_lag1,
            R_pv2p_2024 = 3/4 * R_pv2p + 1/4 * R_pv2p_lag1) %>%
  mutate(pv2p_2024_margin = R_pv2p_2024 - D_pv2p_2024,
         region = tolower(state))

# Add electoral college votes for map labeling
pv2p_2024_states_margins <- pv2p_2024_states_margins %>%
  mutate(year = 2024,
         winner = ifelse(pv2p_2024_margin > 0, "R", "D")) %>%
  left_join(ec, by = c("state", "year"))

# Plot the projected margin of victory in 2024
base_map <- pv2p_2024_states_margins %>%
  left_join(states_map, by = "region") %>%
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pv2p_2024_margin,
                   text = paste("State:", state,
                                "<br>Margin:", round(abs(pv2p_2024_margin), 2),
                                "<br>Projected Winner:", winner,
                                "<br>Electoral Votes:", electors)),
               color = "black") +
  ggtitle("2024 Presidential Margins Forecast (Simplified Electoral Cycle Model)") + 
  scale_fill_gradient2(high = "firebrick1", 
                       mid = "white", 
                       low = "dodgerblue4", 
                       breaks = c(-50, -25, 0, 25, 50),
                       limits = c(-50, 50),
                       name = "Percentage Point Margin (pp)") +
  maps_theme

# Apply plotly to get hover labels
plotly_map <- ggplotly(base_map, tooltip = "text")
plotly_map

```

States with a small projected margin of victory align with the key battlegrounds in 2024: Arizona, Georgia, North Carolina, Pennsylvania, Wisconsin, Michigan, and Nevada. Examining the projected swing of these states compared to the 2020 election helps reveal in which direction these states are moving. 

```{r}
####----------------------------------------------------------#
#### Extension 3 Continued x Forecast 2024
####----------------------------------------------------------#

# Calculate 2024 pv2p, margin, swing
pv2p_2024_states_swing <- d_pvstate_wide %>%
  filter(year == 2020) %>%
  group_by(state) %>%
  mutate(D_pv2p_2024 = 3/4 * D_pv2p + 1/4 * D_pv2p_lag1,
            R_pv2p_2024 = 3/4 * R_pv2p + 1/4 * R_pv2p_lag1) %>%
  mutate(pv2p_2024_margin = R_pv2p_2024 - D_pv2p_2024,
         pv2p_2024_swing =  D_pv2p_2024 - D_pv2p,
         region = tolower(state))

# Add electoral college votes for map labeling
pv2p_2024_states_swing <- pv2p_2024_states_swing %>%
  mutate(year = 2024,
         winner = ifelse(pv2p_2024_margin > 0, "R", "D")) %>%
  left_join(ec, by = c("state", "year"))

# Plot the swing from 2020
base_map <- pv2p_2024_states_swing %>%
  left_join(states_map, by = "region") %>%
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pv2p_2024_swing,
                   text = paste("State:", state,
                                "<br>Swing:", round(pv2p_2024_swing, 2),
                                "<br>Margin:", round(abs(pv2p_2024_margin), 2),
                                "<br>Projected Winner:", winner,
                                "<br>Electoral Votes:", electors)),
               color = "black") +
  ggtitle("2024 Presidential Swing Forecast (Simplified Electoral Cycle Model)") + 
  scale_fill_gradient2(high = "dodgerblue4", 
                       mid = "white", 
                       low = "firebrick1", 
                       breaks = c(-2, -1, 0, 1, 2),
                       limits = c(-2, 2),
                       name = "Percentage Point Swing (pp)") +
  maps_theme

# Apply plotly to get hover labels
plotly_map <- ggplotly(base_map, tooltip = "text")
plotly_map

```

Most states are projected to swing slightly more Republican in 2024 than in 2020, but this does not indicate Republican victory. In fact, this is consistent with the simplified model, relying 75% on the 2020 results and 25% on the 2016 results in each state. Many states voted more Republican in 2016 than they did in 2020, automatically pulling the modeled swings to the Republicans.

```{r, results = 'hide'}
####----------------------------------------------------------#
#### Electoral college 2024 predication
####----------------------------------------------------------#

# calculate 2024 pv2p
pv2p_2024_states_elec <- d_pvstate_wide %>%
  filter(year == 2020) %>%
  group_by(state) %>%
  summarize(D_pv2p_2024 = 3/4 * D_pv2p + 1/4 * D_pv2p_lag1,
            R_pv2p_2024 = 3/4 * R_pv2p + 1/4 * R_pv2p_lag1) %>%
  mutate(region = tolower(state))

# figure out who wins each year, join in state electoral counts
pv2p_2024_states_elec <- pv2p_2024_states_elec %>%
  mutate(year = 2024,
         winner = ifelse(R_pv2p_2024 > D_pv2p_2024, "R", "D")) %>%
  left_join(ec, by = c("state", "year"))

# coutn electors for each party
pv2p_2024_states_elec %>%
  group_by(winner) %>%
  summarize(electoral_votes = sum(electors))

## Harris: 276
## Trump: 262
```

In fact, the model forecasts that Vice President Kamala Harris will win 276 electors to former president Donald Trump's 262, a democratic victory. 

### Simplified Model Pitfalls and Future Forecast Additions
While the simplified model offers a basic prediction, it still misses many of the known drivers of voter behavior. A comparison to [Fair's Electoral Models](https://drive.google.com/drive/u/0/folders/1Sim1ht01NGVFkiCzcx7cOVZo97H02xKj) reveals the absence of  economic fundamentals. The variables in Fair's model are all economy-based, instead using the intercept to account for non-economic values like past voting patterns. 

Additionally, work by Westwood, Messing, and Lelkes (2020) on the [impact of forecasts](https://www-journals-uchicago-edu.ezp-prod1.hul.harvard.edu/doi/pdfplus/10.1086%2F708682) and Enos and Hersch (2015) on the [ineffectiveness of campaign on-the-ground strategy](https://www-cambridge-org.ezp-prod1.hul.harvard.edu/core/journals/american-political-science-review/article/party-activists-as-campaign-advertisers-the-ground-campaign-as-a-principalagent-problem/6117E8EF4C72A046ABFE6A8FDEDEA95E) are not considered.

In future forecasts, I will explore these and other factors that are known to influence elections. For now, the simplified Norpoth model provides a basic 2024 election outcome prediction.

**Current Forecast: Harris 276 - Trump 262**

### Data Sources
- Popular vote by candidate, 1948-2020
- Popular vote by state, 1948-2020
- Electoral College by state, 1948-2020