---
title: 'Week 3: Advanced Regressions for Polling and Fundamentals Models'
author: ''
date: '2024-09-21'
slug: week-3-mutlivariate-analyses
categories: []
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, results = 'hide'}
# Load libraries
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(tidyverse)
library(knitr)
library(kableExtra)
library(plotly)

```

```{r}
####----------------------------------------------------------#
#### Read, merge, and process data
####----------------------------------------------------------#

# FiveThirtyEight polling average datasets national
d_pollav_natl <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/national_polls_1968-2024.csv") # if you run this code please change to location of data

# FiveThirtyEight polling average datasets state
d_pollav_state <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/state_polls_1968-2024.csv") # if you run this code please change to location of data

# pollster rating from 538
d_pollster_ratings <- read_csv("C:/Users/Jacqui Schlesinger/Documents/election-blog1/pollster-ratings-combined.csv")
```
_Please note, this week's blog will be a little more context heavy as I work to establish the building blocks of a model I hope to add onto each week going forward._

This week, I implement new methodologies to build a more complex election forecast model. I begin by examining the differing forecasting methodologies of experienced forecasters, noting areas of potential integration into my own model. Then, looking at the impact of weekly polling data on election outcomes, I apply new regression techniques to better understand their fit as predictors across election years. Finally, I also integrate economic fundamentals into a combined model that will form the basis of my forecasting approach going forward.

### Current Forecasting Methodologies
Many forecasters' models today are based off of a combination of polling data and economic fundamentals. Two forecasters, G Elliot Morris of 538 and Nate Silver of the Silver Bulletin, have published advanced breakdowns of their modeling techniques. I will review the elements of each, noting potential methods that I have implemented here or could do so in the future. 

G Elliott Morris’s [538 forecast model](https://abcnews.go.com/538/538s-2024-presidential-election-forecast-works/story?id=110867585) for the 2024 presidential election employs a sophisticated approach that incorporates polling averages and state correlations through a geometric decay function, emphasizing the interplay of state similarities based on voting behavior, demographics, and geography. The model combines this polling data with a range of economic fundamentals such as employment metrics, consumer sentiment, and inflation to inform its predictions. This combination of fundamentals and polling, with some weighting scheme within and between the two, interests me for further investigation. 538 utilizes Bayesian regression with regularization techniques to manage uncertainty and biases inherent in polling, simulating numerous electoral scenarios through Markov chain Monte Carlo methods-- another future area of work in my model, especially on the simulations front. 

In contrast, Nate Silver's 2020 [(538)](https://fivethirtyeight.com/features/how-fivethirtyeights-2020-presidential-forecast-works-and-whats-different-because-of-covid-19/) and 2024 [(Silver Bulletin)](https://www.natesilver.net/p/model-methodology-2024) models adopt a more dynamic framework that adjusts for contemporary uncertainties, particularly those arising from the COVID-19 pandemic in the 2020 model and changes in voter turnout patterns. Silver was not a fan of The Economists' model when Morris worked there, and additionally [criticized the 538 model](https://www.natesilver.net/p/why-i-dont-buy-538s-new-election) predicting a win for Biden post-debate despite changes in polling towards Trump. Silver’s election forecast combines polling data with factors like economic conditions, incumbency, and demographic trends to predict outcomes. It now adjusts for shifts in voter turnout dynamics, reflecting that Democrats currently have more engaged voters, and incorporates Robert F. Kennedy Jr. as a significant third-party candidate, allowing him to impact the race (pre-dropout). It adjusts for polling quality and biases, while also accounting for uncertainty through factors like national shifts and state-level errors. The model simulates thousands of election scenarios, giving more weight to polling closer to Election Day. Recent changes maintain the model's core methodology while improving accuracy based on recent voting patterns. 

Through these examples, my goal is to achieve a model that can combine factors like polling data, incumbency, and economic fundamentals, as well as use Bayesian models, weighting schemes, and simulation to determine winners in each state. 

### Recent Polling Trends and the Use of Polling in Models
Using polls to reliably determine popular judgement dates back to Francis Galton's 1907 article [Vox populi](https://www.nature.com/articles/075450a0.pdf). Today, thousands of polls are conducted each election cycle to understand the state of the race at a given time. But with so many complications in measuring the behavior of voters, it is extremely difficult for a poll to get the "right" answer-- that is, a single poll can't tell us the what the outcome of the election in some number of days will be.

These difficulties of polling are highlighted in Gelman and King's ["Why are american presidential election campaign polls so variable when votes are so predictable?"](https://www-jstor-org.ezp-prod1.hul.harvard.edu/stable/194212?sid=primo) While the results of polls may fluctuate during a campaign due to specific events, the overall election outcome remains largely predictable based on fundamental political factors. Even predictable events like conventions can cause swings in polls, though these do not have a lasting impact. Voters may appear to change their minds during the campaign, they often make their final decisions based on rational, long-standing preferences instead of short term ones often emphasized by the media. The polling swings are temporary and often do not reflect lasting changes in voter intention.

For an example of this, I can look at poll variation over time in the 2020 election, which is similar in variation to other years. "Game-changers" are also included, indicating reasons why polling averages may have changed at a certain period in time. 

```{r}
####----------------------------------------------------------#
#### Visualizing poll variation over time: 2020
####----------------------------------------------------------#

# polling average 2020 plus labels, 
polling_plot_2020 <- d_pollav_natl |> 
  filter(year == 2020) |> 
  ggplot(aes(x = poll_date, y = poll_support, color = party)) +
  geom_rect(xmin = as.Date("2020-08-17"), xmax = as.Date("2020-08-20"), ymin = 47.5, ymax = 100, alpha = 0.1, color = NA, fill = "grey") + 
  annotate("text", x = as.Date("2020-08-07"), y = 51.5, label = "DNC", size = 4) +
  geom_rect(xmin = as.Date("2020-08-24"), xmax = as.Date("2020-08-27"), ymin = 0, ymax = 47.2, alpha = 0.1, color = NA, fill = "grey") +
  annotate("text", x = as.Date("2020-09-04"), y = 45, label = "RNC", size = 4) +
  geom_rect(xmin = as.Date("2020-10-02"), xmax = as.Date("2020-10-12"), ymin = 0, ymax = 42.7, alpha = 0.05, color = NA, fill = "grey") +
  
  geom_point(size = 1) + 
  geom_line() + 
  
  geom_segment(x = as.Date("2020-03-12"), xend = as.Date("2020-03-12"), y = 0, yend = 44.8, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-03-12"), y = 42.5, label = "COVID \n Market Crash", size = 3) +
  geom_segment(x = as.Date("2020-04-08"), xend = as.Date("2020-04-08"), y = 49, yend = 100, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-03-25"), y = 51.3, label = "Bernie Ends Run", size = 3) +
  geom_segment(x = as.Date("2020-04-16"), xend = as.Date("2020-04-16"), y = 0, yend = 44, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-04-16"), y = 44.7, label = "22 mil \n Unemployment", size = 3) +
  geom_segment(x = as.Date("2020-05-27"), xend = as.Date("2020-05-27"), y = 0, yend = 43, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-06-05"), y = 44, label = "100k COVID Dead, \n George Floyd", size = 3) +
  
  geom_segment(x = as.Date("2020-07-14"), xend = as.Date("2020-07-14"), y = 0, yend = 50.3, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-06-19"), y = 47.5, label = "Moderna Announces", size = 3) +
  
  geom_segment(x = as.Date("2020-09-29"), xend = as.Date("2020-09-29"), y = 50, yend = 100, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-9-12"), y = 49.5, label = "Pres. Debate", size = 3) +
  geom_segment(x = as.Date("2020-10-07"), xend = as.Date("2020-10-07"), y = 51.7, yend = 100, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-10-17"), y = 50.3, label = "VP Debate", size = 3) +
  geom_segment(x = as.Date("2020-10-22"), xend = as.Date("2020-10-22"), y = 52, yend = 100, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-10-30"), y = 51.5, label = "Pres. Debate", size = 3) +
  annotate("text", x = as.Date("2020-10-15"), y = 43.7, label = "Trump Has COVID", size = 3) +
  geom_segment(x = as.Date("2020-09-18"), xend = as.Date("2020-09-18"), y = 50, yend = 100, linetype = "dashed", alpha = 0.4, color = "grey") +
  annotate("text", x = as.Date("2020-09-03"), y = 51.5, label = "RBG Passes", size = 3) +
  
  scale_x_date(date_labels = "%b %d") + 
  scale_color_manual(values = c("dodgerblue4", "firebrick1")) +
  labs(x = "Date",
       y = "Average Poll Approval", 
       title = "Polling Averages by Date with Game Changers, 2020") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"))

polling_plot_2020

```

The uncertainty and variation in polls compared to the final outcome as described by Gelman and King can be clearly seen here. Looking at 2024 below, there is also a large amount of variation in the polling, making modeling based on weekly polling averages difficult without advanced regression coefficient weighting. 

```{r}
####----------------------------------------------------------#
#### Visualizing poll variation over time: 2024
####----------------------------------------------------------#

# create the 2024 polling plot
polling_plot_2024 <- d_pollav_natl |> 
  filter(year == 2024) |> 
  ggplot(aes(x = poll_date, y = poll_support, color = party)) +
  geom_point(aes(text = paste("Date:", poll_date, "<br>Support:", round(poll_support,3), "%"))) + 
  geom_line(stat= "identity") +
  scale_x_date(breaks = scales::date_breaks("1 month"), date_labels = "%b %d") +
  scale_color_manual(values = c("dodgerblue4", "firebrick1")) +
  labs(x = "Date",
       y = "Average Poll Approval", 
       title = "Polling Averages by Date, 2024") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
      axis.text = element_text(size = 9),
      panel.grid.major = element_line(color = "lightgrey", linetype = "dashed")
  )

# convert to interactive plotly
interactive_plot_2024 <- ggplotly(polling_plot_2024, tooltip = "text")
interactive_plot_2024

```

This polling data includes both democratic candidates for this cycle, President Biden and VP Harris. Whether Biden's polling data can be used without adjustment as a proxy for Harris is an open question I will later seek to answer.

### Pollster Quality Evaluation
Different pollster methodologies and decisions can impact the accuracy of their polls. To account for this, 538 creates pollster ratings called "pollscores" which account for bias and error, and also look at transparency, number of polls, and percent partisan work. Looking at the variation in pollster quality can provide valuable information about the quality of the polling data I work with below and throughout the rest of my models.

```{r}
####----------------------------------------------------------#
#### Pollster rating variance calcs
####----------------------------------------------------------#

# mean and 95% CI functions
mean_ci <- function(x) {
  mean_x <- mean(x, na.rm = TRUE)
  se <- sd(x, na.rm = TRUE) / sqrt(length(x))
  ci <- qt(0.95, df = length(x) - 1) * se
  c(mean = mean_x, lower = mean_x - ci, upper = mean_x + ci)
}

# get CI for variety of variables on pollsters
bias_ci <- mean_ci(d_pollster_ratings$bias_ppm)
error_ci <- mean_ci(d_pollster_ratings$error_ppm)
pollscore_ci <- mean_ci(d_pollster_ratings$POLLSCORE)
numeric_grade_ci <- mean_ci(d_pollster_ratings$numeric_grade)
wtd_avg_transparency_ci <- mean_ci(d_pollster_ratings$wtd_avg_transparency)
percent_partisan_work_ci <- mean_ci(d_pollster_ratings$percent_partisan_work)

# get min and max values for comparison column (ended up not using)
bias_min <- min(d_pollster_ratings$bias_ppm, na.rm = TRUE)
error_min <- min(d_pollster_ratings$error_ppm, na.rm = TRUE)
pollscore_min <- min(d_pollster_ratings$POLLSCORE, na.rm = TRUE)
partisan_work_min <- min(d_pollster_ratings$percent_partisan_work, na.rm = TRUE)
numeric_grade_max <- max(d_pollster_ratings$numeric_grade, na.rm = TRUE)
wtd_avg_transparency_max <- max(d_pollster_ratings$wtd_avg_transparency, na.rm = TRUE)

# put together for table
ci_data <- data.frame(
  Metric = c("Bias PPM", "Error PPM", "Pollscore", "Numeric Grade", "Weighted Avg Transparency", "Percent Partisan Work"),
  
  Mean = c(round(bias_ci["mean"], 3), round(error_ci["mean"], 3), round(pollscore_ci["mean"], 3), paste0(round(numeric_grade_ci["mean"], 3), "/3"), paste0(round(wtd_avg_transparency_ci["mean"], 3), "/10"), paste0(round(percent_partisan_work_ci["mean"] * 100, 2), "%")),
  
  Lower_CI = c(round(bias_ci["lower"], 3), round(error_ci["lower"], 3), round(pollscore_ci["lower"], 3), paste0(round(numeric_grade_ci["lower"], 3), "/3"), paste0(round(wtd_avg_transparency_ci["lower"], 3), "/10"), paste0(round(percent_partisan_work_ci["lower"] * 100, 2), "%")),
  
  Upper_CI = c(round(bias_ci["upper"], 3), round(error_ci["upper"], 3), round(pollscore_ci["upper"], 3), paste0(round(numeric_grade_ci["upper"], 3), "/3"), paste0(round(wtd_avg_transparency_ci["upper"], 3), "/10"), paste0(round(percent_partisan_work_ci["upper"] * 100, 2), "%"))
)

# table styling
kable(ci_data) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

These summary statistics for pollster quality reveal the limits of polling data. The current number one ranked pollster (The New York Times/Siena College) has a pollscore of -1.5, the best possible, no partisan work, and 8.7/10 transparency score. It is worth noting that the Predictive Plus-Minus for pollster's absolute error, or error_ppm, is calculated as $predictive\ error = adjusted\ error * (n / (n + n_{shrinkage})) + (group\ error\ prior) * (n_{shrinkage} / (n + n_{shrinkage}))$, and	Predictive Plus-Minus for pollster bias, or bias_ppm, is calculated as $predictive\ bias = adjusted\ bias * (n / (n + n_{shrinkage})) + (group\ error\ prior) * (n_{shrinkage} / (n + n_{shrinkage}))$, per [538 methodology](https://abcnews.go.com/538/538s-pollster-ratings-work/story?id=105398138), where n is the time-weighted number of polls the pollster has released and n_shrinkage is an integer that represents the effective number of polls' worth of weight to put on the prior.

Through these complicated formulas, it is clear that pollster ratings are not particularly inspiring in accuracy of their poll results, at least according to these ratings. It is worth noting that transparency, which is a part of 538's model, is not necessarily a direct correlate with accuracy of results, though this is difficult to test on a poll by poll basis given the lack of true comparison values.

### A Brief Note on Regression Methodologies
Following in class evaluation of Ridge, LASSO, and elastic net regressions using cross-validation to find the optimal lambda values that minimize the MSE of predictions in week-level training data, I have decided to focus on the elastic net regularization method for this week. I will also use model ensembles to combine fundamentals and polling data such that I can weigh the different elements' impact by time until election.

### Polling-Based Regression Models
- build a model (or ensemble model) that uses individual polls from 2016 (president_polls_2016.csv),
2020 (...2020.csv), and 2024 (...2024.csv). How does your model
compare to the models this week in lab?
GRAPH x2

### Combined Fundamentals and Polling Elastic Net Regression Model
- do it 
- in future state by state
REGRESSION TABLE/ EVAL RESULTS

### Data Sources
- Polling average datasets national, state (FiveThirtyEight)
- Pollster aggregate ratings (FiveThirtyEight)