---
title: 'Week 3: Mutlivariate Analyses'
author: ''
date: '2024-09-21'
slug: week-3-mutlivariate-analyses
categories: []
tags: []
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, results = 'hide'}
# Load libraries
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(tidyverse)

```

Review of current forecasting methodologies
- silver 2024
- morris 2024
- fundamentals and poll based models

Regression Methodologies
- ridge versus lasso versus elastic net
- ensemble
- combined with polling data for explanations

Pollster Quality Evaluations and Polling Only Regressions
- 2016-2024 pollster ratings from FiveThirtyEight. 
- How much variation is there in pollster quality
- build a model (or ensemble model) that uses individual polls from 2016 (president_polls_2016.csv),
2020 (...2020.csv), and 2024 (...2024.csv). How does your model
compare to the models this week in lab?

Fundamentals and Polling Elastic Net Regression
- do it 

Future Improvements
- state by state 
