---
title: 'Week 3: Advanced Regressions for Polling and Fundamentals Models'
author: ''
date: '2024-09-21'
slug: week-3-mutlivariate-analyses
categories: []
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, results = 'hide'}
# Load libraries
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(tidyverse)

```

This week, I implement new regression methodologies (ridge, LASSO, elastic net) and examine the impact of granular polling data on election outcomes. I also integrate economic fundamentals into a combined model based on the work of G Elliot Morris and Nate Silver. 

### Current Forecasting Methodologies
Elliott Morris’s 538 forecast model for the 2024 presidential election employs a sophisticated approach that incorporates polling averages and state correlations through a geometric decay function, emphasizing the interplay of state similarities based on voting behavior, demographics, and geography. The model combines this polling data with a range of economic indicators—dubbed “fundamentals”—such as employment metrics, consumer sentiment, and inflation to inform its predictions. It utilizes Bayesian regression with regularization techniques to manage uncertainty and biases inherent in polling, simulating numerous electoral scenarios through Markov chain Monte Carlo methods. In contrast, Nate Silver's 2020 and 2024 models adopt a more dynamic framework that adjusts for contemporary uncertainties, particularly those arising from the COVID-19 pandemic and changes in voter turnout patterns. While both models ultimately blend polling data with fundamentals, Silver's methodology emphasizes a more immediate snapshot of polls closer to election day and incorporates additional layers of uncertainty related to demographic shifts and political polarization. Notably, Morris focuses on correlations between states to predict movements, whereas Silver accounts for broader national trends and uncertainties that affect turnout. Both methodologies highlight the complexity of electoral forecasting and the balance between polling data and fundamental economic indicators, yet they diverge in their treatment of state-level correlations versus overarching national trends.

Review of current forecasting methodologies
- silver 2024
- morris 2024
- fundamentals and poll based models

Polling over the past few years

Regression Methodologies
- ridge versus lasso versus elastic net
- ensemble
- combined with polling data for explanations

Pollster Quality Evaluations and Polling Only Regressions
- 2016-2024 pollster ratings from FiveThirtyEight. 
- How much variation is there in pollster quality
- build a model (or ensemble model) that uses individual polls from 2016 (president_polls_2016.csv),
2020 (...2020.csv), and 2024 (...2024.csv). How does your model
compare to the models this week in lab?

Fundamentals and Polling Elastic Net Regression
- do it 

Future Improvements
- state by state 
